\fullpagehumor[Famous Stories About Some Interesting Bugs]{%
\textbf{1. The 500-Mile Email Limit} \\[4pt]
A university sysadmin faced a bizarre bug: emails wouldn't travel more than 500 miles. Boston at 420 miles worked fine; Memphis at 520 miles failed completely.\\
The cause? A timeout was set in nanoseconds instead of milliseconds. Light travels 200,000 km/s through fiber — the timeout expired at exactly 500 miles.\\[8pt]

\textbf{2. Vanilla Ice Cream vs. Car Security} \\[4pt]
Drivers reported their keyless entry failed — but only after buying vanilla ice cream at one specific shop. Not chocolate, not strawberry. Just vanilla. Engineers were skeptical until they confirmed it.\\
The vanilla counter was at the front, chocolate in back. Vanilla buyers returned before the engine cooled, triggering a temperature-sensitive component failure.\\[8pt]

\textbf{3. Passwords That Depend on Posture} \\[4pt]
An IT team dismissed reports that a password worked standing but failed sitting — until they saw it happen. Same user, same password, different postures, different results.\\
Two keyboard keys had been swapped. Standing users looked down and typed what they saw. Seated users touch-typed from muscle memory, entering the wrong sequence.\\[8pt]

\textbf{4. The Floor Polisher That Crashed Servers} \\[4pt]
Night-shift staff reported server crashes that never occurred during the day. No software changes, no power issues — just nightly failures.\\
The cleaning crew's floor polisher vibrated at 60 Hz, matching certain hard drives' resonant frequency. The vibration misaligned read/write heads just enough to crash the system.\\[8pt]

\textbf{5. Sunlight at 2:45 PM} \\[4pt]
One server crashed daily at exactly 2:45 PM. Logs showed thermal throttling, but only at that precise time.\\
Sunlight through a west-facing window hit the poorly ventilated rack at just the right angle. The temperature spike at 2:45 PM pushed the server past its limit.%
}
