\begin{SideNotePage}{
  \textbf{Hilbert’s Hotel with Infinite Buses:} \par
  An infinite number of buses, each carrying infinitely many passengers, arrives at a hotel with countably infinite rooms. To accommodate everyone, each passenger \(n\) on bus \(m\) is assigned to room \((p_n)^m\), where \(p_n\) is the \(n\)th prime number. 

  \vspace{0.5em}
  No two passengers ever collide, since distinct prime powers are never equal: for \(p \ne q\), there are no integers \(k, \ell\) such that \(p^k = q^\ell\).

  \vspace{1.5em}
  \textbf{Banach-Tarski as a Card Game:} \par
  Imagine a deck with cards \(A, B, A', B'\), where you can make any pile of cards, under one rule: $A$ and $A'$ cancel, $B$ and $B'$ cancel (the cards dissapear if they are next to one another). Each card represents a rigid rotation: \(A, A'\) rotate around one axis; \(B, B'\) around another. In the diagram, blue and red are inverses, as are green and yellow.

  \vspace{0.5em}
  Four players each take all reduced sequences (no canceling cards neighbor each other) beginning with a specific card, forming four disjoint point sets: \(S_A, S_{A'}, S_B, S_{B'}\).

  \vspace{0.5em}
  Now prepend an \(A\) to every sequence in \(S_{A'}\), practiaclly rotating the whole set, forming \(A \cdot S_{A'}\). This transforms sequences that began with \(A'\) into sequences that no longer do. For instance, \(A'BA \mapsto BA\), and \(A' \mapsto\) identity. The union \(S_A \cup A \cdot S_{A'}\) now covers all sequences — forming one full copy of the sphere. Similarly, \(S_B \cup B \cdot S_{B'}\) forms a second.

}{01_BanachTarskiParadox/01_ An Axiom of Your Choice.pdf}
\end{SideNotePage}
